Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-02-25 article-models-tt.tex
extended the rules for identity functions

\subsubsection{Substitution is compatible with the empty context}

The following rules assert that substituting by the term $\jterm{\Gamma}{\emptyf}{\emptytm}$
leaves everything unchanged.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\subst{\emptytm}{A}}{A}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\emptytm}{P}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\emptytm}{f}}{f}}.
\end{align}

@@

\subsubsection{The empty type is compatible with substitution}
The fibers of the empty family are the empty families:
\begin{equation}
\inference{\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}}
\end{equation}

@@

\subsection{The rules establishing the categorical properties of type theory}
\subsubsection{The defining property of weakening}
The judgmental equalities we're about to describe assert that substituting a term
in the weakening a thing gives you the thing back. In the case of contexts we get that each fiber
$\subst{x}{\ctxwk{A}{B}}$ is just $B$ and in the case of terms we get 
that $\ctxwk{A}{y}$ is the constant function
mapping everything to $y:B$. Thus, these rules actually establish the weakening
as the weakening. After stating the rules we will describe what it means to
compose context morphisms (terms of weakened contexts).

\subsubsection{Identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
They are determined by their behavior with respect to substitution combined with
weakening.

\begin{align}
&\inference{\jctx{\Gamma}}{\jterm{\Gamma}{\ctxwk{\Gamma}{\Gamma}}{\idfunc[\Gamma]}}\label{ctxIdFuncForm}\\
&\inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\subst{\idfunc[\Gamma]}{\ctxwk{\Gamma}{A}}}{A}}\label{ctxIdFuncPbInvType}\\
&\inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{\idfunc[\Gamma]}{\ctxwk{\Gamma}{x}}}{x}}\label{ctxIdFuncPbInvTerm}\\
&\inference{\jterm{\Delta}{\ctxwk{\Delta}{\Gamma}}{g}}{\jtermeq{\Delta}{\ctxwk{\Delta}{\Gamma}}{\subst{g}{\ctxwk{\Delta}{\idfunc[\Gamma]}}}{g}}\label{ctxIdFuncSubstInvTerm}
\end{align}

The identity morphisms for types are introduced analogously to the identity
morphisms of contexts, the only difference being that we have a rule asserting
that $\subst{x}{\idfunc[A]}\jdeq x$ for every term $x:A$ in context $\Gamma$.

\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}} \\
& \inference{\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}
\end{align}

Using these rules we can derive that there is a judgmental equality $\jfameq{\ctxext{\Gamma}{A}}{\subst{x}{{\idfunc[A]}{B}}}
{\subst{x}{{x}{B}}}$. Note that $\subst{x}{{x}{B}}$ may just be denoted by
$\subst{x,x}{B}$.

The identity functions in combination with weakening give rise to the projections.
We define

\begin{equation*}
\proj A \defeq \ctxwk{A}{\idfunc[\Gamma]}:\ctxwk{A}{{\Gamma}{\Gamma}}.
\end{equation*}

\subsubsection{Extension on terms}\label{extension-on-terms}
When $\jterm{\Gamma}{A}{x}$ and $\jterm{\Gamma}{\subst{x}{P}}{u}$ are terms,
we obtain a term $\jterm{\Gamma}{\ctxext{A}{P}}{\ctxext{x}{u}}$. This operation
of weakening on terms is like the pairing operation and like with the pairing
operation, extension on terms tells what the terms of an extended family are. 
However, it is handled
much stricter than $\smsym$-types, 
for it has a judgmental inverse and it is subject to various compatibility
rules.
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}\quad\jterm{\Gamma}{\subst{x}{P}}{u}}{\jterm{\Gamma}{\ctxext{A}{P}}{\ctxext{x}{u}}}\\
& \inference{\jterm{\Gamma}{\ctxext{A}{P}}{w}}{\jterm{\Gamma}{A}{\pts{w}}}\\
& \inference{\jterm{\Gamma}{\ctxext{A}{P}}{w}}{\jterm{\Gamma}{\subst{\pts{w}}{P}}{\edg{w}}}\\
& \inference{\jterm{\Gamma}{\ctxext{A}{P}}{w}}{\jtermeq{\Gamma}{\ctxext{A}{P}}{\ctxext{\pts{w}}{\edg{w}}}{w}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jterm{\Gamma}{\subst{x}{P}}{u}}{\jtermeq{\Gamma}{A}{\pts{(\ctxext{x}{u})}}{x}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jterm{\Gamma}{\subst{x}{P}}{u}}{\jtermeq{\Gamma}{\subst{x}{P}}{\edg{(\ctxext{x}{u})}}{u}}
\end{align}

@@

\begin{comment}
\subsubsection{Inclusions}
In ordinary type theory, there is a rule
\begin{equation*}
\inference{\jterm{\Gamma}{A}{x}\qquad\jfam{\Gamma,\,a:A}{P(a)}}
  {\Gamma,\,u:P(x/a)\vdash x:A, u:P(x/a)}
\end{equation*}
This rule gives the inclusion of the fiber $\subst{x}{P}$ into $\ctxext{A}{P}$.

Thus, we require
\begin{equation}
\inference{\jterm{\Gamma}{A}{x}\qquad\jfam{\ctxext{\Gamma}{A}}{P}}
  {\jterm{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxwk{\subst{x}{P}}{\ctxext{A}{P}}}{\typefont{in}^P_x}}
\end{equation}

The terms $\typefont{in}^P_x$ have the following behavior:

\begin{align}
& \inference
    {\jfam{\ctxext{{\Gamma}{A}}{P}}{Q}}
    {\jfameq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{\typefont{in}^P_x}{\ctxwk{\subst{x}{P}}{Q}}}{\subst{x}{Q}}}\\
& \inference
    {\jterm{\ctxext{{\Gamma}{A}}{P}}{Q}{g}}
    {\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\subst{x}{Q}}{\subst{\typefont{in}^P_x}{\ctxwk{\subst{x}{P}}{g}}}{\subst{x}{g}}}
\end{align}
\end{comment}

>>>>

\subsubsection{Substitution is compatible with the empty context}
The fibers of the empty family are the empty families:
\begin{equation}
\inference{\jterm{\Gamma}{A}{x}}{\jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}}
\end{equation}

The following rules assert that substituting by the term $\jterm{\Gamma}{\emptyf}{\emptytm}$
leaves everything unchanged.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jfameq{\Gamma}{\subst{\emptytm}{A}}{A}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{\emptytm}{x}}{x}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\emptytm}{P}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\emptytm}{f}}{f}}.
\end{align}

@@

@@

\subsection{Extension on terms}\label{extension-on-terms}
When $\jterm{\Gamma}{A}{x}$ and $\jterm{\Gamma}{\subst{x}{P}}{u}$ are terms,
we obtain a term $\jterm{\Gamma}{\ctxext{A}{P}}{\ctxext{x}{u}}$. This operation
of weakening on terms is like the pairing operation and like with the pairing
operation, extension on terms tells what the terms of an extended family are. 
However, it is handled
much stricter than $\smsym$-types, 
for it has a judgmental inverse and it is subject to various compatibility
rules.
\begin{align}
& \inference{\jterm{\Gamma}{A}{x}\quad\jterm{\Gamma}{\subst{x}{P}}{u}}{\jterm{\Gamma}{\ctxext{A}{P}}{\ctxext{x}{u}}}\\
& \inference{\jterm{\Gamma}{\ctxext{A}{P}}{w}}{\jterm{\Gamma}{A}{\pts{w}}}\\
& \inference{\jterm{\Gamma}{\ctxext{A}{P}}{w}}{\jterm{\Gamma}{\subst{\pts{w}}{P}}{\edg{w}}}\\
& \inference{\jterm{\Gamma}{\ctxext{A}{P}}{w}}{\jtermeq{\Gamma}{\ctxext{A}{P}}{\ctxext{\pts{w}}{\edg{w}}}{w}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jterm{\Gamma}{\subst{x}{P}}{u}}{\jtermeq{\Gamma}{A}{\pts{(\ctxext{x}{u})}}{x}}\\
& \inference{\jterm{\Gamma}{A}{x}\quad\jterm{\Gamma}{\subst{x}{P}}{u}}{\jtermeq{\Gamma}{\subst{x}{P}}{\edg{(\ctxext{x}{u})}}{u}}
\end{align}

@@

\subsection{Composition and identity functions}
\subsubsection{The defining property of weakening}
The judgmental equalities we're about to describe assert that substituting a term
in the weakening a thing gives you the thing back. In the case of contexts we get that each fiber
$\subst{x}{\ctxwk{A}{B}}$ is just $B$ and in the case of terms we get 
that $\ctxwk{A}{y}$ is the constant function
mapping everything to $y:B$. Thus, these rules actually establish the weakening
as the weakening. After stating the rules we will describe what it means to
compose context morphisms (terms of weakened contexts).

\subsubsection{Identity functions}
Without a rule explicitly asserting the existence of an identity morphism we don't
get one, hence we do that here. The identity morphism is a term which introduced
in ordinary type theory via the variable rule. The variable rule is a bit more
general: it asserts that
\begin{equation*}
\jterm{\Gamma,\,x_1:A_1,\ldots,\,x_n:A_n}{A_i}{x_i}
\end{equation*}
for every $1\leq i\leq n$. Thus, it establishes the projections. In our setting,
we get the projections from the identity morphisms together with weakening. We
already have weakening, so here it suffices to introduce the identity morphisms.
They are determined by their behavior with respect to substitution combined with
weakening. The identity functions will also be subject to compatibility rules.
\begin{align}
& \inference{\jfam{\Gamma}{A}}{\jterm{\ctxext{\Gamma}{A}}{\ctxwk{A}{A}}{\idfunc[A]}}\\
& \inference{\jterm{\Gamma}{A}{x}}{\jtermeq{\Gamma}{A}{\subst{x}{\idfunc[A]}}{x}}\\
& \inference{\jfam{\ctxext{\Gamma}{A}}{P}}{\jfameq{\ctxext{\Gamma}{A}}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}{P}}\\
& \inference{\jterm{\ctxext{\Gamma}{A}}{P}{f}}{\jtermeq{\ctxext{\Gamma}{A}}{P}{\subst{\idfunc[A]}{\ctxwk{A}{f}}}{f}} \\
& \inference{\jterm{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{g}}{\jtermeq{\ctxext{\Gamma}{B}}{\ctxwk{B}{A}}{\subst{g}{\ctxwk{B}{\idfunc[A]}}}{g}}
\end{align}

Using these rules we can derive that there is a judgmental equality 
$\jfameq{\Gamma}{\subst{x}{{\idfunc[A]}{R}}}
{\subst{x}{{x}{R}}}$ for any $\jfam{\ctxext{{\Gamma}{A}}{\ctxwk{A}{A}}}{R}$. 
Note that $\subst{x}{{x}{R}}$ is judgmentatlly the same as
$\subst{\ctxext{x}{x}}{R}$.

The identity functions in combination with weakening give rise to the projections.
We define the term
\begin{equation*}
\proj P \defeq \ctxwk{P}{\idfunc[A]}:\ctxwk{P}{{A}{A}}
\end{equation*}
in context $\ctxext{{\Gamma}{A}}{P}$.

The identity function of an extended family is the extension of the identity
functions on the components in the apropriate way. To find out what the
apropriate way is, note that
\begin{align*}
\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}} & \jdeq \ctxwk{P}{{A}{\ctxext{A}{P}}}\\
& \jdeq \ctxext{\ctxwk{P}{{A}{A}}}{\ctxwk{P}{{A}{P}}}
\end{align*}
We have the term $\jterm{\ctxext{\Gamma}{{A}{P}}}{\ctxwk{P}{A}}{\ctxwk{P}{\idfunc[A]}}$.
Thus we need to find out what $\subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{P}{{A}{P}}}$ is:
\begin{align*}
\subst{\ctxwk{P}{\idfunc[A]}}{\ctxwk{P}{{A}{P}}} & \jdeq \ctxwk{P}{\subst{\idfunc[A]}{\ctxwk{A}{P}}}\\
& \jdeq \ctxwk{P}{P},
\end{align*}
where we find the term $\idfunc[P]$. Therefore, we have the following inference rule:
\begin{equation}
\inference{\jfam{\Gamma}{A}\quad\jfam{\ctxext{\Gamma}{A}}{P}}{\jtermeq{\ctxext{\Gamma}{{A}{P}}}{\ctxwk{\ctxext{A}{P}}{\ctxext{A}{P}}}{\idfunc[\ctxext{A}{P}]}{\ctxext{\ctxwk{P}{\idfunc[A]}}{\idfunc[P]}}}
\end{equation}

The identity function of a weakened family is the weakened identity function:
\begin{equation}
\inference{\jfam{\Gamma}{A}\quad\jfam{\Gamma}{B}}{\jtermeq{\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk{{A}{B}}{{A}{B}}}{\idfunc[\ctxwk{A}{B}]}{\ctxwk{A}{\idfunc[B]}}}
\end{equation}

The identity function of a substituted family is the substitution of the identity function
\begin{equation}
\inference{\jterm{\Gamma}{A}{x}\quad\jfam{\ctxext{\Gamma}{A}}{P}}{\jtermeq{\ctxext{\Gamma}{\subst{x}{P}}}{\ctxwk{\subst{x}{P}}{\subst{x}{P}}}{\idfunc[\subst{x}{P}]}{\subst{x}{\idfunc[P]}}}
\end{equation}
